Beyond Static Retrieval: Adaptive Method for Improving RAG Performance

### 2025-02-01 Paper/latex/paper_RAG.tex
added proposed slution

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\textwidth]{rag.PNG}
	\caption{RAG system component }\label{fig1}
\end{figure}
As illustrated in Figure 1, the RAG system is built around two main components:the retriever and the generator. The retriever?s role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows:
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

where each key \( k_i \) is linked to a corresponding value \( v_i \) (often, the key and value are the same). When a query \( q \) is provided, the goal is to find the top-\( k \) keys that are most similar to the query using a similarity function \( s \), and then retrieve their associated values.

@@

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response \cite{gupta2024comprehensivesurveyretrievalaugmentedgeneration}.

@@

\section{proposed Slution}\label{sec3}

\section{Experimental Results}\label{sec3}
\subsection{Experiments settings}\label{subsec2}

Equations in \LaTeX\ can either be inline or on-a-line by itself (``display equations''). For
inline equations use the \verb+$...$+ commands. E.g.: The equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

For display equations (with auto generated equation numbers)
one can use the equation or align environments:
% KAN Archecture
%\begin{equation}
%f(x_1, x_2, \cdots, x_n)=\sum_{q=1}^{2n+1}\Phi_q %\left( \sum_{p=1}^{n}\phi_{q, p}(x_p)\right)
%\end{equation}

\begin{equation}
\Phi = \{\phi_{p, q}\}, \qquad p=1,2,\cdots, n_{in},\qquad q=1,2,\cdots, n_{out}
\end{equation}

\begin{equation}
x_{l+1,j}=\sum_{i=1}^{n_l}\phi_{l, j, i}(x_{l,i}), \qquad j=1,2,\cdots n_{l+1}
\end{equation}

\begin{equation}
\mathbf{x}_{l+1} = \begin{pmatrix}
\phi_{l,1,1}(\cdot)& \phi_{l,1,2}(\cdot) &\cdots &\phi_{l,1,n_l}(\cdot)  \\
\phi_{l,2,1}(\cdot)&\phi_{l,2,2}(\cdot)  & \cdots &  \phi_{l,2,n_l}(\cdot)\\
\vdots&\vdots  &\ddots  &\vdots  \\
\phi_{l,n_{l+1},1}(\cdot)& \phi_{l,n_{l+1},2}(\cdot) &\cdots  & \phi_{l,n_{l+1},n_l}(\cdot)
\end{pmatrix}\mathbf{x}_{l} =\mathbf{ \Phi}_{l}\mathbf{x}_{l}
\end{equation}
\begin{equation}
\mathbf{KAN(x)}=(\mathbf{\Phi}_{L-1}\circ \mathbf{\Phi}_{L-2}\circ \cdots \circ \mathbf{\Phi}_{1}\circ \mathbf{\Phi}_{0})\mathbf{x}
\end{equation}
% models equations
% system 2
\begin{equation}
y_p(k+1) = f\big[y_p(k), y_p(k-1)\big] + u(k)
\end{equation}
where
\begin{equation*}
f\big[y_p(k), y_p(k-1)\big]=\dfrac{y_p(k)y_p(k-1)[y_p(k)+2.5]}{1+y_p^2(k)+y_p^2(k-1)}
\end{equation*}

% system 3
\begin{equation}
y_p(k+1) = \dfrac{y_p(k)}{1+y_p^2(k)}+ u^3(k)
\end{equation}
% system 4
\begin{equation}
y_p(k+1) = f\big[y_p(k), y_p(k-1), y_p(k-2), u(k), u(k-1)\big] 
\end{equation}
where
\begin{equation*}
f\big[x_1, x_2, x_3,x_4, x_5\big]=\dfrac{x_1x_2x_3x_5(x_3-1)+x_4}{1+x_2^2+x_3^2}
\end{equation*}

@@

\section{Figures}\label{sec6}

\begin{figure}[h]
\centering
\includegraphics[width=0.9\textwidth]{fig.eps}
\caption{This is a widefig. This is an example of long caption this is an example of long caption  this is an example of long caption this is an example of long caption}\label{fig1}
\end{figure}

In case of double column layout, the above format puts figure captions/images to single column width. To get spanned images, we need to provide \verb+\begin{figure*}+ \verb+...+ \verb+\end{figure*}+.

For sample purpose, we have included the width of images in the optional argument of \verb+\includegraphics+ tag. Please ignore this.

\section{Algorithms, Program codes and Listings}\label{sec7}

You may refer above listed package documentations for more details before setting \verb+algorithm+ environment. For program codes, the ``verbatim'' package is required and the command to be used is \verb+\begin{verbatim}+ \verb+...+ \verb+\end{verbatim}+.

Similarly, for \verb+listings+, use the \verb+listings+ package. \verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ is used to set environments similar to \verb+verbatim+ environment. Refer to the \verb+lstlisting+ package documentation for more details.

@@

\begin{algorithm}
\caption{Calculate $y = x^n$}\label{algo1}
\begin{algorithmic}[1]
\Require $n \geq 0 \vee x \neq 0$
\Ensure $y = x^n$ 
\State $y \Leftarrow 1$
\If{$n < 0$}\label{algln2}
        \State $X \Leftarrow 1 / x$
        \State $N \Leftarrow -n$
\Else
        \State $X \Leftarrow x$
        \State $N \Leftarrow n$
\EndIf
\While{$N \neq 0$}
        \If{$N$ is even}
            \State $X \Leftarrow X \times X$
            \State $N \Leftarrow N / 2$
        \Else[$N$ is odd]
            \State $y \Leftarrow y \times X$
            \State $N \Leftarrow N - 1$
        \EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

To reference line numbers in an algorithm, consider the label declared for the line number 2 of Algorithm~\ref{algo1} is \verb+\label{algln2}+. To cross-reference it, use the command \verb+\ref{algln2}+ for which it comes up as line~\ref{algln2} of Algorithm~\ref{algo1}.

>>>>

\begin{figure}[h]
	\centering
	\includegraphics[width=0.9\textwidth]{rag.PNG}
	\caption{RAG system component }\label{rag.PNG}
\end{figure}
As illustrated in figure \ref{rag.PNG}, the RAG system is built around two main components:the retriever and the generator. The retriever?s role is to find relevant information from a pre-built data store, while the generator utilizes this retrieved data to produce coherent and contextually appropriate content. The overall RAG process operates as follows:
\subsection{Retrieval in RAG Systems}\label{subsec2}

@@

where each key \( k_i \) is linked to a corresponding value \( v_i \) (often, the key and value are the same). When a query \( q \) is provided, the goal is to find the top-\( k \) keys that are most similar to the query using a similarity function \( s \), and then retrieve their associated values\cite{gupta2024comprehensivesurveyretrievalaugmentedgeneration}.

@@

\subsection{Generator  in RAG Systems}\label{subsec2}
In Retrieval-Augmented Generation (RAG) systems, the generator is a key component responsible for crafting the final output. It works by combining the retrieved information with the users input query to produce a well-structured and contextually relevant response .

@@

\section{proposed Slution}\label{sec3}
In traditional candidate selection systems, the reliance on static or fixed top-K retrieval lead to suboptimal retrieval performance, especially in scenarios where the relevance distribution of candidates is highly variable. To address these limitations, we propose Dynamic Candidate Selection, a novel algorithm that adaptively refines the candidate selection process by leveraging component-level embeddings and a Mixture of Logits (MoL) scoring mechanism.

The \textbf{Mixture of Logits (MoL)}\cite{zhai2023revisiting} methode provides a flexible and adaptive mechanism for computing similarity scores between queries and items. Given a query \( q \) and an item \( x \), MoL assumes that both are mapped to \( P \) groups of low-rank embeddings, denoted as \( f_p(q) \) and \( g_p(x) \), respectively. These embeddings are parameterized by neural networks that extract features from the query and item. The similarity between \( q \) and \( x \) is computed as a weighted sum of the inner products of these embeddings, where the weights \( \pi_p(q, x) \)\cite{zhai2023revisiting} are adaptive gating parameters constrained to sum to 1:
\[
\text{Similarity}(q, x) = \sum_{p=1}^P \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle.
\]
This formulation allows MoL to capture complex relationships between queries and items, making it particularly suitable for dynamic candidate selection tasks.

Inspired by the approach in\cite{ding2024efficient}, we introduce an adaptive threshold mechanism (\textbf{Dynamic Candidate Selection}), the MoL framework is employed to refine the candidate retrieval process. Specifically:
\begin{enumerate}
	\item \textbf{Component-Level Embeddings:}  
	 generate component-level embeddings for all items in \( X \). These embeddings allow efficient similarity computations during retrieval.
	\begin{equation*}
		X_p \leftarrow \{ g_p(x) \mid x \in X \}
	\end{equation*}

	\item \textbf{Dynamic Threshold Adjustment:}  
	To refine retrieval quality, we compute Mixture of Logits (MoL) scores for each candidate \( x \in G \).  
	The adaptive gating weights \( \pi_p(q, x) \) enable the algorithm to dynamically adjust the retrieval threshold \( T_{\text{adaptive}} \) based on the computed MoL scores.
	\begin{equation*}
		\varphi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
	\end{equation*}
	\begin{equation*}
		T_{\text{adaptive}} = \min \{ s \mid s \in G \}
	\end{equation*}

	\item \textbf{Refinement and Top-K Selection:}  
	Additional relevant candidates are retrieved using the adaptive threshold \( T_{\text{adaptive}} \), expanding the candidate set \( G' \). The algorithm then selects the most relevant top-k results by sorting \( G' \) based on MoL scores.
	\begin{equation*}
		G' \leftarrow G \cup \{ x \mid s_p \geq T_{\text{adaptive}} \}
	\end{equation*}
	Finally, we sort candidates by MoL scores and extract the exact top-k:
	\begin{equation*}
		G_{\text{final}} = \text{Top-}k(G', \varphi(q, x))
	\end{equation*}
\end{enumerate}
the flowchart provided in Figure~\ref{fig2} describes the step-by-step process of the proposed algorithm. It visualizes the key stages of the solution
\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\textwidth]{Flowchart_Steps.PNG}
	\caption{Algorithm workflow. }\label{fig2}
\end{figure}

As shown in Algorithm~\ref{Algo1}, the retrieval process dynamically adjusts the threshold based on MoL scores.
\begin{algorithm}
	\caption{Hybrid Exact Top-k with Threshold-Based k Selection}\label{ALgo1}
	\begin{algorithmic}[1]
		\State \textbf{Input:}
		\begin{itemize}
			\item Query $q$
			\item Set of items $X$
			\item Component-level embeddings: $f_p(q)$, $g_p(x)$ for $p \in P$, $x \in X$
			\item Initial threshold $T_{\text{init}}$
		\end{itemize}

		\State \textbf{Output:}
		\begin{itemize}
			\item Exact top $k$ items based on dynamic threshold selection, $G_{\text{final}}$
		\end{itemize}

		\State \textbf{1. Initialize:}
		\begin{itemize}
			\item Set $G \gets \emptyset$ \Comment{Initial candidate set}
		\end{itemize}

		\State \textbf{2. Generate Component-Level Embeddings:}
		\For{each component $p \in P$}
		\State $X_p \gets \{g_p(x) \mid x \in X\}$ \Comment{Precompute embeddings}
		\EndFor

		\State \textbf{3. Initial Candidate Retrieval:}
		\For{each component $p \in P$}
		\State Compute dot product scores: 
		\[
		S_p = \{ \langle f_p(q), g_p(x) \rangle : x \in X_p \}
		\]
		\State Retrieve items with scores $S_p \geq T_{\text{init}}$
		\State Add these items to $G$
		\EndFor

		\State \textbf{4. Adjust k Dynamically:}
		\For{each $x \in G$}
		\State Compute MoL scores $s = \phi(q, x)$ for each $x \in G$ using:
		\[
		\phi(q, x) = \sum_{p=1}^{P} \pi_p(q, x) \cdot \langle f_p(q), g_p(x) \rangle
		\]
		\State Set $T_{\text{adaptive}} = \min \{s : s \in G\}$
		\EndFor

		\State \textbf{5. Refine Candidate Set with Adaptive k:}
		\For{each component $p \in P$}
		\State Retrieve items from $X_p$ with scores $S_p \geq T_{\text{adaptive}}$
		\State Add these items to $G'$
		\EndFor

		\State \textbf{6. Select Exact Top-k Items:}
		\State Compute MoL scores for all items in $G'$
		\State Sort $G'$ by MoL scores in descending order
		\State Select the top $k$ items from $G'$ where $k$ is the number of items in $G'$ exceeding $T_{\text{adaptive}}$

		\State \textbf{7. Return:} $G_{\text{final}}$ \Comment{Top $k$ items from $G'$}
	\end{algorithmic}
\end{algorithm}
\section{Experimental Results}\label{sec3}
\subsection{Experiments settings}\label{subsec2}

@@

@@

@@
